---
title: "HPAclusteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HPAclusteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`HPAclusteR` is an R package designed by the Human Protein Atlas to streamline the process of gene clustering from transcriptomics data. It provides a modular pipeline for performing PCA, calculating distances, constructing shared nearest neighbor (SNN) graphs, consensus clustering, UMAP, functional annotation, and different publication-ready visualizations. The package is designed to work with [AnnDatR objects](https://github.com/emiliosk/AnnDatR), making it easy to integrate into workflows for transcriptomics data analysis.

>💡 This vignette demonstrates the step-by-step usage of the `HPAclusteR` pipeline. For users who prefer simplicity, the `hc_auto_cluster()` function can be used to run the entire pipeline in one call.

## Getting Started

To begin, load the `HPAclusteR` package:

```{r setup}
library(HPAclusteR)
```

We will use the built-in `example_adata` dataset for this vignette. This dataset is a toy example designed to demonstrate the functionality of the package.

## Step 1: Principal Component Analysis (PCA)

PCA is the first step in the pipeline, used to reduce the dimensionality of the data while retaining the most important features.

```{r pca, message=FALSE, warnings=FALSE}
adata_res <- hc_pca(example_adata, components = 40)
head(adata_res$obsm$X_pca[, 1:5])  # PCA scores
adata_res$uns$pca  # Explained variance
```

The PCA step reduces the dimensionality of the data to 40 components, which can be adjusted based on the dataset and analysis goals.

## Step 2: Distance Calculation

Next, we calculate the pairwise distances between samples using the PCA-reduced data. The `hc_distance()` function supports multiple distance metrics, such as "euclidean" and "spearman". We will use the number of components determined by Kaiser’s rule.

```{r distance, message=FALSE, warnings=FALSE}
adata_res <- hc_distance(
  adata_res, 
  components = hc_kaisers_rule(adata_res), 
  method = "spearman"
)
head(adata_res$uns$distance)
```

## Step 3: Shared Nearest Neighbor (SNN) Graph Construction

The SNN graph is constructed to identify clusters of similar samples. This step uses the distance matrix calculated in the previous step.

```{r snn, message=FALSE, warnings=FALSE}
adata_res <- hc_snn(adata_res, neighbors = 15, prune = 1 / 15)
adata_res$uns$neighbors$snn
```

The `neighbors` parameter controls the number of nearest neighbors considered, while `prune` adjusts the sparsity of the graph.

## Step 4: Consensus Clustering

Consensus clustering is performed to identify robust clusters in the data. This step aggregates clustering results from multiple runs, in this case 100.

```{r consensus, message=FALSE, warnings=FALSE}
adata_res <- hc_cluster_consensus(
  adata_res, 
  resolution = 6, 
  method = "louvain",
  n_seeds = 100
)

head(adata_res$obs$cluster)
```

The `resolution` parameter controls the granularity of the clustering. Higher values result in more clusters.

## Step 5: UMAP Visualization

UMAP is used to visualize the clusters in a low-dimensional space. This step provides an intuitive way to explore the clustering results. It uses the constructed SNN graph.

```{r umap, message=FALSE, warnings=FALSE}
adata_res <- hc_umap(adata_res, verbose = FALSE)
head(adata_res$obsm$X_umap)
```

## Step 6: Cluster Hulls (Optional)

Cluster hulls are calculated to visualize the boundaries of each cluster in the UMAP plot.

```{r hulls, message=FALSE, warnings=FALSE}
adata_res <- hc_cluster_hulls(adata_res, poly_smoothing = 4, relative_bandwidth = 1/40)
head(adata_res$uns$UMAP_hulls$hulls)
```

## Step 7: Visualization

Finally, we visualize the UMAP plot with clusters and hulls using the `hc_plot_umap()` function.

```{r plot, message=FALSE, warnings=FALSE}
hc_plot_umap(adata_res, plot = "both")
```

The `plot` argument can be set to `"points"`, `"hulls"`, or `"both"` to customize the visualization.

## Conclusion

Explore the package further to uncover the full potential of your transcriptomics data!